---
title: "Lab 9: Multivalued and Continuous Instrumental Variables"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Learning objectives

By the end of this lab, you should be able to:

1. Explain why the simple binary-IV LATE story does not transfer mechanically to multivalued and continuous instruments.
2. Estimate and interpret a 2SLS model with a multivalued instrument.
3. Estimate and interpret a 2SLS model with a continuous instrument.
4. Use binary thresholding as a practical approximation when you want Wald-style intuition or complier profiling.
5. Propose a feasible extension of an existing IV paper for a final replication project.

# Why this lab?

Lab 8 focused on the cleanest IV design: a binary instrument, a binary treatment, and a clear complier group. That setup is useful because the Wald estimator has a direct Local Average Treatment Effect (LATE) interpretation.

This week we move one step further. In many empirical applications:

- the instrument has more than two values, or
- the instrument varies continuously, or
- the treatment itself is not binary.

Once that happens, the algebra of IV still works, but the interpretation becomes less simple. A 2SLS coefficient is usually a weighted average of different local responses rather than a single binary-treatment LATE.

To keep the intuition from Lab 8 alive, we will combine two ideas:

1. estimate the main multivalued or continuous IV model directly; and
2. build binary approximations from substantively meaningful thresholds when we want Wald intuition or complier profiling with `ivdesc`.

This makes the lab useful both for understanding the theory and for thinking about realistic final replication projects.

# Packages

## Install packages if needed

```{r install-packages, eval=FALSE}
install.packages(c("dplyr", "ggplot2", "ivreg", "ivdesc", "knitr", "micsr"))
```

## Load packages

```{r load-packages}
library(dplyr)
library(ggplot2)
library(ivreg)
library(ivdesc)
library(knitr)
library(micsr)

theme_set(theme_minimal())
```

## Helper functions

```{r helper-functions}
extract_term <- function(model, term, label) {
  s <- coef(summary(model))
  tibble(
    model = label,
    term = term,
    estimate = unname(s[term, "Estimate"]),
    std.error = unname(s[term, "Std. Error"])
  )
}

profile_binary_iv <- function(data, var, D, Z, label = var) {
  out <- as.data.frame(
    ivdesc::ivdesc(
      X = data[[var]],
      D = data[[D]],
      Z = data[[Z]],
      boot = FALSE
    )
  )

  out %>%
    mutate(
      group = factor(
        c("Sample", "Complier", "Never-taker", "Always-taker"),
        levels = c("Sample", "Complier", "Never-taker", "Always-taker")
      ),
      variable = label,
      lower = mu - 1.96 * mu_se,
      upper = mu + 1.96 * mu_se
    ) %>%
    select(variable, group, mu, mu_se, lower, upper)
}
```

# 1. From binary IV to multivalued and continuous IV

The binary IV story from Lab 8 is still the benchmark:

- binary `Z`,
- binary `D`,
- Wald = reduced form / first stage,
- interpretation = LATE for one complier group.

When `Z` is not binary, two important things change.

## Multivalued instrument

If `Z` has several values, 2SLS uses several instrument-induced contrasts at once. Instead of one first-stage coefficient, we now have a *set* of first-stage shifts.

The practical interpretation is:

- 2SLS combines several binary comparisons inside one regression;
- the estimate is a weighted average of local responses induced by those shifts;
- there is no single universal complier group unless we impose an extra binary approximation.

## Continuous instrument

If `Z` is continuous, the same idea goes further:

- the first stage is a slope, not a difference in means;
- the reduced form is also a slope;
- the just-identified IV estimator is still "reduced form divided by first stage", but the interpretation is no longer "the effect for one complier group."

Following Angrist and Imbens' variable-treatment-intensity logic, the coefficient is better understood as a weighted average of marginal causal responses among units whose treatment is shifted by the instrument.

## Why we will still use binary approximations

Packages such as `ivdesc` are built for the binary-IV, binary-treatment case. That means complier profiling is easiest when we define:

- a binary threshold for the instrument, and
- a binary threshold for treatment uptake.

This is not the same estimand as the original multivalued or continuous IV design. It is a pedagogical and exploratory approximation that helps us understand *who* is moved by a particular threshold contrast.

# 2. Example 1: Multivalued IV with college proximity

## The empirical case

Our first paper is **Card (1995), "Using Geographical Variation in College Proximity to Estimate the Return to Schooling."**

The key idea is that growing up close to different types of colleges shifts the cost of attending college. We use the `SchoolingReturns` teaching dataset distributed with the `ivreg` package, which is based on Card's study.

This is a good multivalued-IV example because:

- the instrument has more than two values (`nearcollege4 = none, public, private`);
- the treatment (`education`) is continuous;
- the dataset is bundled in a CRAN package, so it is reproducible without any external download.

## Main variables

- `nearcollege4`: grew up near no 4-year college, a public 4-year college, or a private 4-year college (multivalued instrument)
- `education`: years of completed education (endogenous treatment)
- `wage`: hourly wage in 1976 (outcome)
- `experience`, `ethnicity`, `smsa`, `south`, `age`: baseline controls used in standard teaching replications

## Load the teaching dataset

```{r card-data}
data("SchoolingReturns", package = "ivreg")

card <- SchoolingReturns %>%
  mutate(
    nearcollege4 = factor(nearcollege4, levels = c("none", "public", "private")),
    log_wage = log(wage)
  )
```

## Descriptive summaries

```{r card-college-summary}
card %>%
  group_by(nearcollege4) %>%
  summarise(
    n = n(),
    mean_education = mean(education),
    mean_wage = mean(wage)
  ) %>%
  kable(digits = 3)
```

```{r card-first-stage-plot}
card_means <- card %>%
  group_by(nearcollege4) %>%
  summarise(
    mean_education = mean(education),
    mean_log_wage = mean(log_wage),
    .groups = "drop"
  )

ggplot(card_means, aes(x = nearcollege4, y = mean_education, fill = nearcollege4)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  labs(
    title = "Average education by college proximity",
    subtitle = "College proximity shifts years of schooling",
    x = NULL,
    y = "Average years of education"
  )
```

## OLS versus multivalued 2SLS

The multivalued IV regression uses `nearcollege4` as a factor instrument. With factor coding, one category is omitted as the baseline and the remaining instrument contrasts are used jointly in 2SLS.

```{r card-models}
m_ols_card <- lm(
  log_wage ~ education + poly(experience, 2, raw = TRUE) + ethnicity + smsa + south,
  data = card
)

m_first_card <- lm(
  education ~ nearcollege4 + poly(age, 2, raw = TRUE) + ethnicity + smsa + south,
  data = card
)

m_first_restricted_card <- lm(
  education ~ poly(age, 2, raw = TRUE) + ethnicity + smsa + south,
  data = card
)

m_iv_card <- ivreg(
  log_wage ~ education + poly(experience, 2, raw = TRUE) + ethnicity + smsa + south |
    nearcollege4 + poly(age, 2, raw = TRUE) + ethnicity + smsa + south,
  data = card
)
```

```{r card-first-stage-test}
anova(m_first_restricted_card, m_first_card)
```

```{r card-iv-summary}
summary(m_iv_card, diagnostics = TRUE)
```

```{r card-compare}
bind_rows(
  extract_term(m_ols_card, "education", "OLS"),
  extract_term(m_iv_card, "education", "2SLS with multivalued college-proximity instrument")
) %>%
  select(model, estimate, std.error) %>%
  kable(digits = 3)
```

## Interpretation

This is the key lesson:

- there is **no single Wald ratio** here because `nearcollege4` is not binary;
- instead, 2SLS combines multiple proximity contrasts at once;
- the resulting coefficient should be read as a weighted average of education effects induced by college-proximity variation.

## Approximation: recover a binary contrast for complier profiling

To get back to the logic of Lab 8, we can build an exploratory binary approximation:

- define a binary instrument `near4 = 1` for respondents who grew up near any 4-year college;
- define a binary treatment `some_college = 1` for at least 13 years of schooling.

This creates a *threshold-specific* encouragement design. It is not the same estimand as the full multivalued-IV model, but it gives us an interpretable approximation for profiling compliers.

```{r card-profile-data}
card_profile <- card %>%
  mutate(
    near4 = as.integer(nearcollege4 != "none"),
    some_college = as.integer(education >= 13),
    smsa_yes = as.integer(smsa == "yes"),
    south_yes = as.integer(south == "yes"),
    afam = as.integer(ethnicity == "afam"),
    library14_yes = as.integer(library14 == "yes")
  )

card_profile %>%
  summarise(
    n = n(),
    near_college_share = mean(near4),
    some_college_share = mean(some_college)
  ) %>%
  kable(digits = 3)
```

```{r card-profile-plot}
card_profiles <- bind_rows(
  profile_binary_iv(card_profile, "feducation", "some_college", "near4", "Father's education"),
  profile_binary_iv(card_profile, "meducation", "some_college", "near4", "Mother's education"),
  profile_binary_iv(card_profile, "library14_yes", "some_college", "near4", "Library card at 14")
)

ggplot(card_profiles, aes(x = mu, y = group)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0, linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(
    title = "Approximate complier profiles for a college-proximity threshold",
    subtitle = "These profiles use a binary approximation, not the full multivalued-IV estimand",
    x = "Mean and 95% confidence interval",
    y = NULL
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Discussion

1. Why is the multivalued 2SLS estimate not identical to any single binary Wald comparison?
2. What exactly changes when we collapse `nearcollege4` into a binary college-proximity indicator?
3. Do the approximate complier profiles suggest that the threshold-specific compliers look different from the sample overall?

## Possible extension for a final replication project

One useful extension is to study **which type of college proximity matters most**.

For example:

- compare public-college proximity to private-college proximity directly;
- test heterogeneity by family background, race, or region;
- compare the full multivalued-IV estimate with several binary approximations (`any 4-year college` vs `none`, `public` vs `none`, `private` vs `none`) to show how the implied complier group changes.

# 3. Example 2: Continuous IV with cigarette taxes

## The empirical case

Our second paper uses the birth-weight application from **Mullahy (1997), "Instrumental-Variable Estimation of Count Data Models: Applications to Models of Cigarette Smoking Behavior."**

The dataset in `micsr::birthwt` contains:

- `birthwt`: infant birth weight;
- `cigarettes`: number of cigarettes smoked per day during pregnancy;
- `cigtax`: per-pack state excise tax on cigarettes.

This is a useful continuous-IV example because:

- the treatment is continuous (`cigarettes`);
- the instrument is continuous (`cigtax`);
- the data are easy to access in R;
- the topic makes the difference between OLS and IV substantively meaningful.

## Load and inspect the data

```{r bw-data}
data("birthwt", package = "micsr")

bw <- birthwt %>%
  mutate(
    any_smoke = as.integer(cigarettes > 0),
    high_tax = as.integer(cigtax >= median(cigtax))
  )

bw %>%
  summarise(
    n = n(),
    mean_birthwt = mean(birthwt),
    mean_cigarettes = mean(cigarettes),
    mean_cigtax = mean(cigtax)
  ) %>%
  kable(digits = 3)
```

```{r bw-scatter}
ggplot(bw, aes(x = cigtax, y = cigarettes)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  labs(
    title = "Continuous first stage: cigarette tax and smoking",
    x = "Cigarette tax",
    y = "Cigarettes per day during pregnancy"
  )
```

## OLS, reduced form, first stage, and IV

We include a common set of controls:

- `parity`
- `race`
- `sex`
- `edmother`
- `edfather`
- `faminc`

```{r bw-models}
m_ols_bw <- lm(
  birthwt ~ cigarettes + parity + race + sex + edmother + edfather + faminc,
  data = bw
)

m_first_bw <- lm(
  cigarettes ~ cigtax + parity + race + sex + edmother + edfather + faminc,
  data = bw
)

m_rf_bw <- lm(
  birthwt ~ cigtax + parity + race + sex + edmother + edfather + faminc,
  data = bw
)

m_iv_bw <- ivreg(
  birthwt ~ cigarettes + parity + race + sex + edmother + edfather + faminc |
    cigtax + parity + race + sex + edmother + edfather + faminc,
  data = bw
)
```

```{r bw-first-stage}
summary(m_first_bw)
```

```{r bw-iv-summary}
summary(m_iv_bw, diagnostics = TRUE)
```

```{r bw-wald}
wald_bw <- coef(m_rf_bw)["cigtax"] / coef(m_first_bw)["cigtax"]
wald_bw
```

```{r bw-compare}
bind_rows(
  extract_term(m_ols_bw, "cigarettes", "OLS"),
  extract_term(m_iv_bw, "cigarettes", "2SLS"),
  tibble(
    model = "Reduced form divided by first stage",
    term = "cigarettes",
    estimate = unname(wald_bw),
    std.error = NA_real_
  )
) %>%
  select(model, estimate, std.error) %>%
  kable(digits = 3)
```

## Interpretation

The ratio of reduced form to first stage still reproduces the just-identified IV estimate, but the interpretation is different from the binary-IV case:

- OLS compares mothers with different smoking behavior directly and is likely confounded;
- IV uses only the smoking variation shifted by cigarette taxes;
- the coefficient is best understood as a weighted average response among **tax-sensitive smoking margins**, not as a single binary-treatment LATE for one neat complier group.

## Approximation: threshold-specific complier profiling

Again, we can recover Lab-8-style intuition by defining:

- `high_tax = 1` for states above the sample median tax;
- `any_smoke = 1` for mothers who smoke at all.

That gives a binary threshold contrast. It is an approximation to the original continuous-IV design, but it lets us use `ivdesc` to describe the mothers whose smoking behavior is shifted by this threshold.

```{r bw-profile-plot}
bw_profiles <- bind_rows(
  profile_binary_iv(bw, "faminc", "any_smoke", "high_tax", "Family income"),
  profile_binary_iv(bw, "edmother", "any_smoke", "high_tax", "Mother's education"),
  profile_binary_iv(bw, "parity", "any_smoke", "high_tax", "Parity")
)

ggplot(bw_profiles, aes(x = mu, y = group)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0, linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(
    title = "Approximate complier profiles for a high-tax threshold",
    subtitle = "This is a binary approximation to a continuous-instrument design",
    x = "Mean and 95% confidence interval",
    y = NULL
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Discussion

1. Why does the ratio-of-slopes algebra still work when `Z` and `D` are continuous?
2. Why does the interpretation of the IV estimate change even though the algebra looks similar?
3. How sensitive do you think the complier profile would be to using a different tax threshold?

## Possible extension for a final replication project

A strong extension here is to study **which margin of smoking responds to taxes**.

For example:

- compare the extensive margin (`any_smoke`) to the intensive margin (`cigarettes` among smokers only);
- test whether the IV estimate differs by maternal education or race;
- move closer to Mullahy's original contribution by comparing linear 2SLS to a nonlinear count-model specification.

# 4. Final takeaway

The main message from this lab is that the **estimation mechanics of IV generalize more easily than the interpretation**.

- With a binary instrument, LATE and complier profiling are direct.
- With a multivalued instrument, 2SLS aggregates several local shifts.
- With a continuous instrument, 2SLS aggregates marginal responses along the treatment margin affected by the instrument.
- Binary thresholding is often a useful approximation when you want to recover Wald intuition or profile threshold-specific compliers.

That is why multivalued and continuous IV designs are best approached in two steps:

1. estimate the main IV model honestly; and
2. use binary approximations carefully, and transparently, when they help you understand who is being moved by the instrument.
